---
title: 'Reproducible Research: Peer Assessment 1'
author: "Steve Chiu"
date: "12 April 2016"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

The zip file "activity.zip", loaded in the working directory, was unzipped and read into *activity*.  An *actclean* dataset was also created to ignore the *NA* cases:

```{r, echo=TRUE}
unzip("activity.zip")
activity <- read.csv("activity.csv", header = TRUE)
head(activity)
str(activity)
actclean <- activity[complete.cases(activity),]
head(actclean)
```

## What is mean total number of steps taken per day?

From *actclean* we generate the total number of steps taken per day into *daysteps*, followed by cleaning up those dates with no record, by calling the *camplete.cases* function:

```{r, echo=TRUE}
daysteps <- with(actclean, tapply(steps,date,sum))
daysteps <- daysteps[complete.cases(daysteps)]
head(daysteps)
```

Then we calculate the mean of the daily total steps from *daysteps*:

```{r, echo=TRUE}
print("Mean = ")
mean(daysteps)
```

A histogram was saved as *plot1.png*, with the mean marked in red line.

```{r, echo=TRUE}
hist(daysteps, xlab = "Total Number of Steps Taken Per Day", main = "Histogram")
abline(v=mean(daysteps), col = "red", lwd = 2)

print("Medium = ")
median(daysteps)
```

## What is the average daily activity pattern?

First we generate the total number of steps taken per interval into *timesteps*.  A line chart was created in *plot2.png*.

```{r, echo=TRUE}
timesteps <- with(actclean, tapply(steps, interval, mean))
# Transformed into a dataframe for the time series plotting
meansteps <- data.frame(as.numeric(names(timesteps)), timesteps)
names(meansteps) = c("interval", "steps")

plot(meansteps, type="l", xlab="5-min Intervals", ylab="Average Number of Steps Across All Days", main="Average Daily Activity Pattern")
```

The maximum of steps can be provided by the *max* and *which.max* functions:

```{r, echo=TRUE}
max(meansteps$steps)
which.max(meansteps$steps)
```

The average daily activity reaches the maximum of 37.94 at the 104th time interval, which starts at 835 min.

## Imputing missing values

Going back to the original dataset *activity*, number of records with NA is:

```{r, echo=TRUE}
sum(is.na(activity$steps))
```

To fill up the missing values, we use the average number of steps (per each time interval) across all the dates, as shown in the *steps* column of the new dataset *actnew* below:

```{r, echo=TRUE}
library(dplyr)
activity %>%
      arrange(interval, date) %>%
      group_by(interval) %>%
      mutate(activity$steps <- ifelse(is.na(steps), mean(steps, na.rm=TRUE), steps)) -> actnew
actnew <- arrange(as.data.frame(actnew[,2:4]), date, interval)
names(actnew)[3] <- "steps"
head(actnew)
```

We apply the calculation of the total number of steps taken per day into *daystepsnew*:

```{r, echo=TRUE}
daystepsnew <- with(actnew, tapply(steps,date,sum))
head(daystepsnew)
```

Then we calculate the mean of the daily total steps from *daystepsnew*:

```{r, echo=TRUE}
print("Mean = ")
mean(daystepsnew)
```

A histogram was created in *plot3.png*, with the mean marked as green line.

```{r, echo=TRUE}
hist(daystepsnew, xlab = "New Total Number of Steps Taken Per Day", main = "Histogram")
abline(v=mean(daystepsnew), col = "green", lwd = 2)

print("Medium = ")
median(daystepsnew)
```

It seems that, after filling in the missing values, the mean and medium are very close (or identitical) to the original ones, and the overall impact is minimal.

## Are there differences in activity patterns between weekdays and weekends?

We create two new columnes, namely *wd* and *wde* to show the day of the week and weekday/weekend respectively:

```{r, echo=TRUE}
wd <- weekdays(as.Date(actnew$date, "%Y-%m-%d"))
actnew <- cbind(actnew, wd)
actnew$wde <- sapply(actnew$wd, FUN = function(x) {if (x=="Saturday"| x=="Sunday") x <- "weekend" else x <- "weekday"})
actnew$wde <- as.factor(actnew$wde)
```

Two line charts are drawn below to compare the daily activity pattern between weekdays and weekends (saved as plot4.png).

```{r, echo=TRUE}
library(ggplot2)
actwd <- subset(actnew, wde=="weekday")
actwe <- subset(actnew, wde=="weekend")
timestepswd <- with(actwd, tapply(steps, interval, mean))
timestepswe <- with(actwe, tapply(steps, interval, mean))
meanstepswd <- data.frame(as.numeric(names(timestepswd)), timestepswd)
meanstepswe <- data.frame(as.numeric(names(timestepswe)), timestepswe)
names(meanstepswd) = c("interval", "steps")
names(meanstepswe) = c("interval", "steps")
meanstepswd <- cbind(meanstepswd, wde="weekday")
meanstepswe <- cbind(meanstepswe, wde="weekend")
meanstepswde <- rbind(meanstepswe, meanstepswd)

g <- ggplot(meanstepswde, aes(interval, steps))
g + geom_line() + facet_grid(wde ~ .) + labs(x="Interval") + labs(y="Number of Steps")
```
